---
title: "Introduction to AMRdb"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('AMRdb')`"
vignette: >
  %\VignetteIndexEntry{Introduction to AMRdb}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
library(tidyverse)
library(AMRdb)
```

## Installation 

Get the latest stable `R` release from [CRAN](http://cran.r-project.org/). 

```{r 'CRAN_install', eval = FALSE}
install.packages("AMRdb")
```

To install the development version from [Github](https://github.com) use: 

```{r 'github_install', eval = FALSE}
if (!requireNamespace("remotes", quietly = TRUE)) {
    install.packages("remotes")
}

remotes::install_github("EveliaCoss/AMRdb")
```

Or you can install `AMRdb` from [Bioconductor](http://bioconductor.org/) using the following code:

```{r 'install', eval = FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE)) {
    install.packages("BiocManager")
}

BiocManager::install("AMRdb")
```

And the development version from [GitHub](https://github.com/EveliaCoss/AMRdb) with:

```{r 'install_dev', eval = FALSE}
BiocManager::install("EveliaCoss/AMRdb")
```

## AMRdb package

This package contains a curated collection of datasets designed to support antimicrobial resistance (AMR) analysis across public and clinical sources.

- We focus primarily on `antibiograms_db`, a harmonized dataset integrating susceptibility profiles retrieved from ENA, BV-BRC, and NCBI, covering multiple bacterial genera. It enables out-of-the-box exploration of phenotypic resistance across antibiotics and taxa and serves as a robust alternative to ad hoc clinical tables.

- Gene-level predictions are also available via `rgi_results`, a dataset produced using CARD-RGI annotations that assigns ARO-based resistance features to samples. This genomic metadata allows comparative analyses between phenotypic and genotypic resistance trends.

For custom modeling and validation workflows, AMRdb provides cleaned metadata for both test and training subsets:

- `training_db_cleaned` and `test_db_cleaned` include standardized MIC values, genus annotations, and **curated phenotype assignments** after preprocessing and recategorization.

- The corresponding **raw metadata**, stored in `training_metadata_db` and `test_metadata_db`, preserves all original accession-level values and annotations for traceability and auditing.

Together, these resources support reproducible classification, comparative genomics, and resistance prediction pipelines in a user-friendly format.

### Exploring Resistance: A Deep Dive into `antibiograms_db`

The `AMRdb::antibiograms_db` dataset contains standardized antimicrobial susceptibility records across diverse bacterial genera, with 14 variables and over `r length(unique(antibiograms_db$biosample))` biosamples. It integrates phenotype and taxonomy metadata from ENA, NCBI BioSample, and BV-BRC, enabling out-of-the-box analysis of resistance patterns and comparative modeling.

```{r}
str(antibiograms_db)
```
The `AMRdb::antibiograms_db` data contains

```{r}
# Extract unique antibiotic names:
antibiotics <- sort(unique(antibiograms_db$antibiotic))
# Initialize classification table:
antibiotic_class <- data.frame(
  antibiotic = antibiotics,
  drug_class = NA_character_,  # Puedes usar factor si prefieres
  stringsAsFactors = FALSE
)

# Assign drug class categories:
antibiotic_class <- antibiotic_class %>%
  mutate(drug_class = case_when(
    str_detect(antibiotic, "penicillin|amoxicillin|ampicillin|oxacillin|methicillin|ticarcillin|piperacillin") ~ "Beta-lactam",
    str_detect(antibiotic, "cefa|ceph|cef") ~ "Cephalosporin",
    str_detect(antibiotic, "carbapenem|imipenem|meropenem|doripenem|biapenem") ~ "Carbapenem",
    str_detect(antibiotic, "macrolide|azithromycin|clarithromycin|erythromycin|spiramycin|telithromycin") ~ "Macrolide",
    str_detect(antibiotic, "tetracycline|doxycycline|oxytetracycline|minocycline|chlortetracycline") ~ "Tetracycline",
    str_detect(antibiotic, "aminoglycoside|gentamicin|tobramycin|streptomycin|amikacin|netilmicin|neomycin") ~ "Aminoglycoside",
    str_detect(antibiotic, "sulfa|sulfamethoxazole|sulfisoxazole|sulfadimethoxine|sulfonamide") ~ "Sulfonamide",
    str_detect(antibiotic, "quinolone|ciprofloxacin|levofloxacin|norfloxacin|moxifloxacin|gatifloxacin") ~ "Fluoroquinolone",
    str_detect(antibiotic, "glycopeptide|vancomycin|teicoplanin|dalbavancin") ~ "Glycopeptide",
    TRUE ~ "Other/Unclassified"
  ))

# Merge classifications back into original dataset:
antibiograms_db <- antibiograms_db %>%
  left_join(antibiotic_class, by = "antibiotic")

# Plot 
antibiograms_db %>%
  count(drug_class, sort = TRUE) %>%
  ggplot(aes(x = reorder(drug_class, n), y = n)) +
  geom_col(fill = "darkcyan") +
  coord_flip() +
  labs(title = "Frequency of Drug Classes in AMRdb",
       x = "Drug Class",
       y = "Sample Count") +
   theme_classic()
```






## Citing `AMRdb`

We hope that `r Biocpkg("AMRdb")` will be useful for your research. Please use the following information to cite the package and the overall approach. Thank you!

```{r "citation"}
## Citation info
citation("AMRdb")
```

Here is an example of you can cite your package inside the vignette:

* `r Biocpkg("AMRdb")`
