---
title: "Download raw data from public sources"
output: 
  BiocStyle::html_document:
    self_contained: yes
    toc: true
    toc_float: true
    toc_depth: 2
    code_folding: show
date: "`r doc_date()`"
package: "`r pkg_ver('AMRdb')`"
vignette: >
  %\VignetteIndexEntry{Download raw data from public sources}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}  
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    crop = NULL ## Related to https://stat.ethz.ch/pipermail/bioc-devel/2020-April/016656.html
)
```


```{r vignetteSetup, echo=FALSE, message=FALSE, warning = FALSE}
## Bib setup
library("RefManageR")

## Write bibliography information
bib <- c(
    R = citation(),
    BiocStyle = citation("BiocStyle")[1],
    knitr = citation("knitr")[1],
    RefManageR = citation("RefManageR")[1],
    rmarkdown = citation("rmarkdown")[1],
    sessioninfo = citation("sessioninfo")[1],
    testthat = citation("testthat")[1],
    AMRdb = citation("AMRdb")[1]
)
```

Load packages

```{r packages, message=FALSE, warning=FALSE, eval=FALSE}
library(here)
library((janitor)
library(tidyverse)
```

## STEP1 - Opcion A: Download data from URL

### Combined antibiogram dataset

You can download the antibiogram files from [ENA browser](https://www.ebi.ac.uk/ena/browser/home), [BV-BRC](https://www.bv-brc.org/), and [NCBI](https://www.ncbi.nlm.nih.gov/) using the reference file available on [Zenodo](https://zenodo.org/records/14876710), published on July 4, 2025 (version 4).

```{r, eval=FALSE}
# Define URL and local destination
anti_url <- "https://zenodo.org/records/15809334/files/antibiograms.tsv.zip"
# Replace "inst/extdata" with the actual path to your folder
targetfile <- here("inst/extdata", "antibiograms.tsv.zip") # <- Change this to your own directory

# Download the compressed file
download.file(url = anti_url, destfile = targetfile, mode = "wb")

# Unzip the .zip file
unzip(targetfile, exdir = "inst/extdata")

# Import the data into R
antibiograms_metadata <- readr::read_tsv(here("inst/extdata","antibiograms.tsv"))  # <- Specify the directory path
```

You can compress the file to `.gz` format before loading it into R. The `read_tsv()` function supports reading `.gz-compressed` files directly.

```{r, eval=FALSE}
# Compress the file to .gz format
original_file <- here("inst/extdata", "antibiograms.tsv")
gz_file <- here("inst/extdata", "antibiograms.tsv.gz")
R.utils::gzip(original_file, destname = gz_file, overwrite = TRUE)

# Read the .tsv.gz file into R
antibiograms_metadata <- readr::read_tsv(here("inst/extdata", "antibiograms.tsv.gz"))
```

We recommend simplifying column formats using the `janitor` package and standardizing column names as shown below.

```{r, eval=FALSE}
# Clean antibiogram metadata
antibiograms_db <- antibiograms_metadata %>%
  # Standardizes column names: converts them to lowercase,
  # replaces spaces and special characters with underscores
  janitor::clean_names() %>%
  # Renames the columns
  rename(
    accession = reads,
    scientific_name_Antibiogram = species
  )
```

### CAMDA training dataset

```{r, eval=FALSE}
# Define the raw URL
url <- "https://raw.githubusercontent.com/EveliaCoss/AMRdb/devel/inst/extdata/training_dataset.csv"

# Destination filename
dest <- "training_dataset.csv"

# Download the file
download.file(url, dest, mode = "wb")

# Load the data
training_metadata <- read.csv(dest, stringsAsFactors = FALSE)

# Rename species
training_metadata_db <- training_metadata %>%
  mutate(scientific_name_CAMDA = paste(genus, species, sep = " "))
```


### CAMDA test dataset

```{r, eval=FALSE}
# Define the raw URL
url <- "https://raw.githubusercontent.com/EveliaCoss/AMRdb/devel/inst/extdata/testing_template.csv"

# Destination filename
dest <- "testing_template.csv"

# Download the file
download.file(url, dest, mode = "wb")

# Load the data
test_metadata <- read.csv(dest, stringsAsFactors = FALSE)

# Rename species
test_metadata_db <- test_metadata %>%
  mutate(scientific_name_CAMDA = paste(genus, species, sep = " ")) %>%
  mutate(antibiotic = if_else(antibiotic == "tetracycline", "TET", antibiotic))
```

## STEP1 - Opcion B: Download data from packages

### Combined antibiogram dataset

Download antibiogram files from [ENA browser](https://www.ebi.ac.uk/ena/browser/home), [BV-BRC](https://www.bv-brc.org/), and [NCBI](https://www.ncbi.nlm.nih.gov/) via the [Zenodo](https://zenodo.org/records/14876710) reference (version 4, July 4, 2025).

```{r, eval=FALSE}
library(AMRdb)

dir <- system.file("extdata", package="AMRdb")
# Read the .tsv.gz file into R
antibiograms_metadata <- readr::read_tsv(file.path(dir, "antibiograms.tsv.gz"))
```

### CAMDA training dataset

```{r package_training_download, eval=FALSE}
dir <- system.file("extdata", package="AMRdb")
training_metadata <- readr::read_csv(file.path(dir, "training_dataset.csv"))
# Rename species
training_metadata_db <- training_metadata %>%
  mutate(scientific_name_CAMDA = paste(genus, species, sep = " "))
# Rows: 6144 Columns: 16── Column
```

### CAMDA test dataset

The data were provided by CAMDA as part of the 2025 challenge focused on phenotype and MIC prediction using publicly available datasets from 9 bacterial species. However, we have identified that the test dataset includes duplicated entries or information already present in the training dataset. This redundancy may affect model evaluation and should be carefully examined during preprocessing.

```{r package_test_download, eval=FALSE}
# Load CAMDA test dataset
dir <- system.file("extdata", package="AMRdb")
# Read the .csv file into R
test_metadata <- readr::read_csv(file.path(dir, "testing_template.csv"))

# Rename species
test_metadata_db <- test_metadata %>%
  mutate(scientific_name_CAMDA = paste(genus, species, sep = " ")) %>%
  mutate(antibiotic = if_else(antibiotic == "tetracycline", "TET", antibiotic))
# Rows: 5345 Columns: 6── Column
```

### NCBI information

```{r, eval=FALSE}
# Read the .csv file into R
dir <- system.file("extdata", package="AMRdb")
sra_metadata_db <- readr::read_csv(file.path(dir, "sra-metadata.csv") %>%
  # Standardizes column names: converts them to lowercase,
  # replaces spaces and special characters with underscores
  janitor::clean_names() %>%
  # Select columns
  select(run, library_strategy, library_selection, library_source, platform,
         model, bio_project, bio_sample, scientific_name) %>%
  # Renames columns
  rename(accession = run, scientific_name_complete = scientific_name)

# Reduce name
sra_metadata_db$scientific_name_NCBI <- word(sra_metadata_db$scientific_name_complete, 1, 2)
```

### Genome taxonomy database (GTDB)

The information extracted from the Genome Taxonomy Database (GTDB) is linked to the accession IDs provided in both the CAMDA training and test datasets

```{r, eval=FALSE}
dir <- system.file("extdata", package="AMRdb")
# test information
test_gtdb_result <- readr::read_csv(file.path(dir, "test.ani.csv"))
# training information
train_gtdb_result <- readr::read_csv(file.path(dir, "train.ani.csv"))
```

## Compare to package data

Compare to `?antibiograms_db`:

```{r}
library(AMRdb)
data(package = "AMRdb")

head(antibiograms_db)
```




